---
title: "exploratory"
author: "Kelly Sovacool"
date: "2020 Apr 03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Random Forest model

Useful links:

- caret docs: https://topepo.github.io/caret/measuring-performance.html#measures-for-predicted-classes
- Geoff's paper: https://mbio.asm.org/content/9/6/e02248-18
- Geoff's code: https://github.com/SchlossLab/Hannigan_CRCVirome_mBio_2018/blob/master/bin/predictVirome.R
- Begüm's code: https://github.com/SchlossLab/Topcuoglu_ML_XXX_2019/blob/master/code/learning/individual_models/RandomForest.R
- confusion matrix: https://en.wikipedia.org/wiki/Confusion_matrix

### First, try on example data
```{r}
library(mlbench)
data(Sonar)
head(Sonar)
```

### Split the data into training & testing
```{r}
library(caret)
set.seed(998)
inTraining <- createDataPartition(Sonar$Class, p = .75, list = FALSE)
training <- Sonar[ inTraining,]
testing  <- Sonar[-inTraining,]
```

### 5-fold Cross-validation
```{r}
cross_val <- trainControl(method="repeatedcv",
                     repeats = 1,
                     number=5,
                     returnResamp="final",
                     classProbs=TRUE,
                     summaryFunction=twoClassSummary,
                     indexFinal=NULL,
                     savePredictions = TRUE)
```

### Training
```{r}
model <- train(Class ~ ., 
               data = training, 
               trControl = cross_val, 
               method="rf", 
               metric="ROC", 
               tuneLength=5)
model
```


### Feature importance
```{r}
rf_imp <- varImp(model)
plot(rf_imp, top=20)
sort(rf_imp$Overall, decreasing = TRUE)
rf_imp_final <- varImp(model$finalModel)
sort(rf_imp_final$Overall, decreasing = TRUE)
```

### Measuring performance
```{r}
#roc_imp <- filterVarImp(x = training[, -ncol(training)], y = training$Class)
#head(roc_imp)
testing$pred = predict(model, newdata = testing)
model_prec_recall <- confusionMatrix(
    data = testing$pred,
    reference = testing$Class,
    mode = "prec_recall"
)
print(model_prec_recall)
model_sens_spec <- confusionMatrix(
    data = testing$pred,
    reference = testing$Class)
print(model_sens_spec)
```

### Now try it on some of our metadata + fake data

```{r}
library(here)
library(tidyverse)
metadata <- read_csv(here('data', 'SraRunTable.txt'))
head(metadata)
```

Eventually, we'll need to merge the metadata with the mothur shared file (OTU abundances).

For now, let's make up some stuff.

First, only use samples that are cancerious or healthy.
```{r}
metadata %>% group_by(DiseaseClass) %>% summarise(n = n())
diagnoses <- metadata %>% filter(DiseaseClass %in% c("Cancer", "Healthy"))
head(diagnoses)
```

Rename sonar data classes to cancer/healthy for fake data for development purposes.
```{r}
otu_abun_data <- Sonar %>% 
    mutate(DiseaseClass = case_when(
        Class == 'R' ~ 'Cancer',
        Class == 'M' ~ 'Healthy'
    )) %>% 
    select(-Class)
head(otu_abun_data)
```

Test our machine learning code:
```{r}
source(here('code', "machine_learning.R"))
set.seed(999)
inTraining <-
    createDataPartition(otu_abun_data %>% pull(DiseaseClass),
                        p = 0.75,
                        list = FALSE)
training_data <- otu_abun_data[inTraining, ]
testing_data  <- otu_abun_data[-inTraining, ]
train_control <- trainControl(
    method = "repeatedcv",
    repeats = 5,
    number = 5,
    returnResamp = "final",
    classProbs = TRUE,
    summaryFunction = twoClassSummary,
    indexFinal = NULL,
    savePredictions = TRUE
)
otu_model <- train(
    DiseaseClass ~ .,
    data = training_data,
    trControl = train_control,
    method = "rf",
    metric = "Mean_AUC",
    tuneLength = 15
)
testing_data <- testing_data %>% 
    mutate(prediction = predict(otu_model, newdata= testing_data),
           DiseaseClass = as.factor(DiseaseClass))
prec_recall <- confusionMatrix(
    data = testing_data$prediction,
    reference = testing_data$DiseaseClass,
    mode = "prec_recall"
)
prec_recall
```

### OTU (bacteria)
```{r}
source(here('code', "machine_learning.R"))
otu_abun <-
    read_tsv(here("data", '16S', 'mothur_output', 'stability.opti_mcc.shared')) %>%
    mutate(
        subject_id = str_to_title(Group) %>%
            str_replace("([A-Z])[a-z]*([0-9]{1,2})_[_0-9]*", "\\1\\2"),
        DiseaseClass = case_when(
            startsWith(Group, 'A') ~ 'Adenoma',
            startsWith(Group, 'C') ~ 'Cancer',
            startsWith(Group, 'H') ~ 'Healthy'
        )
    ) %>%
    filter(DiseaseClass %in% c("Cancer", "Healthy")) %>% 
    select(-label, -Group, -numOtus, -subject_id) %>%
    select(DiseaseClass, everything())
head(otu_abun)
set.seed(545)
inTraining <- as.vector(
    createDataPartition(otu_abun %>% pull(DiseaseClass),
                        p = 0.6,
                        list = FALSE
                        )
    )
training_data <- otu_abun[inTraining, ]
testing_data  <- otu_abun[-inTraining, ]
train_control <- trainControl(
    method = "repeatedcv",
    repeats = 5,
    number = 5,
    returnResamp = "final",
    classProbs = TRUE,
    summaryFunction = twoClassSummary,
    indexFinal = NULL,
    savePredictions = TRUE
)
otu_model <- train(
    DiseaseClass ~ .,
    data = training_data,
    trControl = train_control,
    method = "rf",
    metric = "Mean_AUC",
    tuneLength = 15
)
testing_data <- testing_data %>% 
    mutate(prediction = predict(otu_model, newdata= testing_data),
           DiseaseClass = as.factor(DiseaseClass))
prec_recall <- confusionMatrix(
    data = testing_data$prediction,
    reference = testing_data$DiseaseClass,
    mode = "prec_recall"
)
```

### OVU (viruses)
```{r}
bins <- read_csv(here("data","virome","concoct","clustering_merged.csv"), col_types = 'cc') %>% 
    rename(contig = contig_id)
coverage <- read_tsv(here("data","virome","contigs","coverage_table.tsv")) %>% 
    rename_at(vars(contains("SRR")), funs(str_replace(., ".*(SRR.*)_mapped.sorted", "\\1"))) %>% 
    mutate(contig = str_replace(contig, "([0-9]*)\\..*", "\\1")) %>% 
    pivot_longer(-contig, names_to = "Run", values_to = "abun")
metadata <- read_csv(here('data', 'SraRunTable.txt')) %>% 
    select(subjectid, Run, DiseaseClass)
ovu_abun <- coverage %>% 
    inner_join(metadata, by = "Run") %>%  
    inner_join(bins, by = "contig") %>% 
    select(subjectid, DiseaseClass, contig, abun, cluster_id) %>% 
    mutate(cluster_id = str_replace(cluster_id, "(.*)", "OVU\\1")) %>% 
    group_by(subjectid, cluster_id, DiseaseClass) %>% 
    summarize(abunsum = sum(abun)) %>% 
    pivot_wider(names_from = cluster_id, values_from = abunsum)
head(ovu_abun)
```

### Plot AUROC
```{r plot_roc}
library(plotROC)
library(RColorBrewer)
color_disease <- brewer.pal(3, "Dark2")
color_seq <- brewer.pal(3, "Paired")
otu <- read_tsv(here("data", "model", "rf_model_bacteria.tsv")) %>% 
    mutate(seqtype = "16S Bacteria")
ovu <- read_tsv(here("data", "model", "rf_model_virus.tsv")) %>% 
    mutate(seqtype = "Virus")
models <- rbind(otu, ovu)
ggplot(models, aes(d = obs, 
                   m = Healthy, 
                   color = seqtype)) + 
    geom_roc(n.cuts=0) +
    style_roc(theme = theme_gray()) +
    coord_equal() +
    #theme_classic() +
    scale_color_manual(values = c("#1F78B4", "#B2DF8A")) +
	theme(
		legend.position = c(0.75, 0.25),
		legend.title=element_blank()
	) +
    ggtitle("Random Forest Model Performance")
```

### Feature importance

```{r}
feat_imp_virus <- readRDS(here("data", "model", "feat_imp_virus.rds"))
feat_imp_bact <- readRDS(here("data", "model", "feat_imp_bacteria.rds"))
feat_imp_bact$otu = rownames(feat_imp_bact)
feat_imp_bact <- feat_imp_bact %>% 
    as_tibble() %>% 
    arrange(desc(Overall)) %>% 
    rename(importance = Overall)
# TODO: link OTU to tax
```

