---
title: "exploratory"
author: "Kelly Sovacool"
date: "2020 Apr 03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Random Forest model

Useful links:

- caret docs: https://topepo.github.io/caret/measuring-performance.html#measures-for-predicted-classes
- Geoff's paper: https://mbio.asm.org/content/9/6/e02248-18
- Geoff's code: https://github.com/SchlossLab/Hannigan_CRCVirome_mBio_2018/blob/master/bin/predictVirome.R
- Begüm's code: https://github.com/SchlossLab/Topcuoglu_ML_XXX_2019/blob/master/code/learning/individual_models/RandomForest.R
- confusion matrix: https://en.wikipedia.org/wiki/Confusion_matrix

### First, try on example data
```{r}
library(mlbench)
data(Sonar)
head(Sonar)
```

### Split the data into training & testing
```{r}
library(caret)
set.seed(998)
inTraining <- createDataPartition(Sonar$Class, p = .75, list = FALSE)
training <- Sonar[ inTraining,]
testing  <- Sonar[-inTraining,]
```

### 5-fold Cross-validation
```{r}
cross_val <- trainControl(method="repeatedcv",
                     repeats = 1,
                     number=5,
                     returnResamp="final",
                     classProbs=TRUE,
                     summaryFunction=twoClassSummary,
                     indexFinal=NULL,
                     savePredictions = TRUE)
```

### Training
```{r}
model <- train(Class ~ ., 
               data = training, 
               trControl = cross_val, 
               method="rf", 
               metric="ROC", 
               tuneLength=5)
model
```


### Feature importance
```{r}
rf_imp <- varImp(model)
plot(rf_imp, top=20)
sort(rf_imp$Overall, decreasing = TRUE)
rf_imp_final <- varImp(model$finalModel)
sort(rf_imp_final$Overall, decreasing = TRUE)
```

### Measuring performance
```{r}
#roc_imp <- filterVarImp(x = training[, -ncol(training)], y = training$Class)
#head(roc_imp)
testing$pred = predict(model, newdata = testing)
model_prec_recall <- confusionMatrix(
    data = testing$pred,
    reference = testing$Class,
    mode = "prec_recall"
)
print(model_prec_recall)
model_sens_spec <- confusionMatrix(
    data = testing$pred,
    reference = testing$Class)
print(model_sens_spec)
```

### Now try it on some of our metadata + fake data

```{r}
library(here)
library(tidyverse)
metadata <- read_csv(here('data', 'SraRunTable.txt'))
head(metadata)
```

Eventually, we'll need to merge the metadata with the mothur shared file (OTU abundances).

For now, let's make up some stuff.

First, only use samples that are cancerious or healthy.
```{r}
metadata %>% group_by(DiseaseClass) %>% summarise(n = n())
diagnoses <- metadata %>% filter(DiseaseClass %in% c("Cancer", "Healthy"))
head(diagnoses)
```

Rename sonar data classes to cancer/healthy for fake data for development purposes.
```{r}
otu_abun_data <- Sonar %>% 
    mutate(DiseaseClass = case_when(
        Class == 'R' ~ 'Cancer',
        Class == 'M' ~ 'Healthy'
    )) %>% 
    select(-Class)
head(otu_abun_data)
```

Test our machine learning code:
```{r}
source(here('code', "machine_learning.R"))
set.seed(999)
inTraining <-
    createDataPartition(otu_abun_data %>% pull(DiseaseClass),
                        p = 0.75,
                        list = FALSE)
training_data <- otu_abun_data[inTraining, ]
testing_data  <- otu_abun_data[-inTraining, ]
otu_model <- run_rf(training_data)
testing_data <- testing_data %>% 
    mutate(prediction = predict(otu_model, newdata= testing_data),
           DiseaseClass = as.factor(DiseaseClass))
prec_recall <- confusionMatrix(
    data = testing_data$prediction,
    reference = testing_data$DiseaseClass,
    mode = "prec_recall"
)
prec_recall
```

### OTU (bacteria)
```{r}
otu_abun <- read_tsv(here("data", '16S', 'mothur_output', 'stability.opti_mcc.shared'))
head(otu_abun)
otu_abun %>% mutate(subject_id = 
                        str_to_title(Group) %>% 
                        str_replace("([A-Z])[a-z]*([0-9]{1,2})_[_0-9]*", 
                                    "\\1\\2"),
                    DiseaseClass = case_when(
                        startsWith(Group, 'A') ~ 'Adenoma',
                        startsWith(Group, 'C') ~ 'Cancer',
                        startsWith(Group, 'H') ~ 'Healthy'
                    )
                    ) %>% 
    pull(subject_id)
```

### OVU (viruses)
