---
title: "Investigating the microbial community of the human gut in colorectal cancer"
author: "Nicole Bowers, Brittany Hicks, Christina Kang-Yun, Katelyn Polemi, Kelly Sovacool"
csl: mbio.csl
output:
  pdf_document:
    includes:
      in_header: preamble_report.tex
    keep_tex: no
    latex_engine: xelatex
    toc: yes
  word_document:
    toc: yes
  html_document:
    df_print: paged
    toc: yes
geometry: margin=1in
link-citations: yes
fontsize: 11pt
bibliography: refs_report.bib
subparagraph: true # fixed error that says "! Argument of \paragraph has an extra }."
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE,
                      message = FALSE,
                      fig.path = here::here("figures/")
                      )
```
```{r load}
library(broom)
library(here)
library(kableExtra)
library(knitr)
library(tidyverse)
```

\newpage
\todo{Guidelines: https://drive.google.com/file/d/1YNqYiMJYiVWXiSR9yWUbQTPAoUpVa9Id/view}

# Abstract

# Introduction

# Methods

## Study design and data collection

\todo{study design / data collection based on hannigan and zackular papers}
[@hannigan_diagnostic_2017; @Zackular_human_2014]

The data are available in the NCBI Sequence Read Archive as [PRJNA389927](https://www.ncbi.nlm.nih.gov/bioproject/389927).
All code used in preparation of this report is available in the following GitHub repository: https://github.com/kelly-sovacool/bioinf545-group3-project.

## 16S rRNA gene sequence processing
The 16S rRNA gene sequences associated with this study were previously reported by [@Zackular_human_2014]Sequence (fastq) and metadata files were downloaded from the following site: http://www.mothur.org/MicrobiomeBiomarkerCRC.
The 16S rRNA gene sequences were analyzed with the mothur software package (v1.37.0) [@kozich_Development_2013] [@schloss_introducing_2009]. Mothur allows contigs from forward and reverse fastq files to be generated, pre-processing steps for primer removal and quality screening, and the identification of unique sequences. After these steps were taken to reduce sequencing and PCR errors, sequences were aligned to the SILVA database [@pruesse_SILVA_2007], screened for chimeras using VSEARCH, and binned into operational taxonomic units (OTUs) using a 97% similarity threshold. 


## Metagenome and virome quality control

Trimmomatic (v.0.39) [@bolger_trimmomatic_2014] was used to remove adapter sequences and low-quality reads in metagenome and virome sample sets. 
The read quality was assessed using FastQC (v.0.11.9) [@andrews_fastqc_2010] before and after adapter trimming to confirm removal of adapters and low quality reads.
Unpaired reads were dropped using the repair function in BBTools (v.37.62) [@bushnell_bbtools_2018].
Then, reads mapping to the human GRCh38 reference genome were removed using the BWA-MEM algorithm (BWA v.0.7.12) [@li_aligning_2013].
The paired unmapped reads were used for taxonomic profiling and gene annotation.

## Virome assembly
Closely related viral contig sequences were categorized into operational viral units (OVUs) in the absence of taxonomic identity. OVUs were defined with the CONCOCT algorithm (v0.4.0). This algorithm applies a variation-based Bayesian approach that bins related contigs by similar tetramer and coabundance profiles within samples [@alneberg_binning_2014].


## Classification modeling

## Metagenome taxonomic profiling and gene annotation

The human genome-free reads were used to profile the metagenomic taxonomy at the species, genus, family, and phylum levels using MetaPhlAn2 [@truong_metaphlan2_2015].
The results were visualized using the R package phyloseq [@mcmurdie_phyloseq_2013].

The paired reads that did not map to the human genome were aligned to the Integrated Gene Catalog (IGC) [@li_integrated_2014] of 1,267 gut microbiome samples consisting of approximately 10 million genes with the BWA MEM algorithm for metagenome annotation [@li_aligning_2013]. 
Annotated genes were extracted from the alignment results using functions geneList and countKegg modified from the MGS-Fast pipeline [@brown_mgs-fast_2019]. 
The differences in gene abundance between healthy, adenoma, and cancer groups were assessed using the R package edgeR [@robinson_edger_2009]. 
Up to 480 top KEGG numbers of genes that were significantly different between healthy and other groups were selected and the relevant KEGG pathway was determined using the KEGG mapper tool [@kanehisa_kegg_2020].

# Results

## Bacterial diversity in colorectal cancer

The influence of colorectal cancer on bacterial OTU diversity was evaluated amongst the three disease classes. A total of 3,904 OTUs were identified. Alpha diversity was evaluated based upon Shannon diversity metrics. Shannon’s diversity index accounts for both the abundance and the evenness of the species present. As seen in Figure XX, individuals with healthy colons do not appear to have a distinctly different gut microbiome from an evenness perspective. The figure suggests that individuals with cancerous and healthy colons may have a select group of OTUs that dominate the environment since their index values are not relatively high. There was no statistically significant clustering of the disease groups (analysis of variance [ANOVA] P value = 0.3).

The beta diversity was evaluated based upon Bray-Curtis metrics. Bray-Curtis dissimilarity allows for the calculation of statistical dissimilarity between the three disease classes relative to each other. As seen in Figure XX, there does not appear to be distinct clustering of the disease classes. Instead, each disease class is fairly dispersed. There was no statistically significant clustering of the disease groups ([ANOVA] P value > 0.23). 


## Taxnomonic abundance based on metagenome



## Metagenome annotation and quantification

\begin{wrapfigure}{R}{0.5\textwidth}
    \begin{center}
        \includegraphics[width=0.49\textwidth]{../figures/gene_abundance_MDS.pdf}  
        \caption{Multidimensional scaling (MDS) abundance plot based on KEGG annotation counts derived from gut microbiome of patients that are healthy (H) or diagnosed with adenoma (A) or cancer (C).}
        \label{fig:MDS_KeggNo}  
    \end{center}
\end{wrapfigure}

The gut metagenome from patients that are healthy or diagnosed with adenoma or cancer was aligned to the IGC database to annotate the genes with KEGG numbers.
KEGG number counts were calculated for all samples, then were used to determine whether specific genes were more or less abundant in patients with adenoma or cancer as compared to those who are healthy using edgeR.
Seven samples, including the negative control, with extremely low read counts were excluded from subsequent analyses, as they strongly skewed the results.
Multidimensional scaling (MDS) analysis of the KEGG number counts between different groups did not show clear clustering within groups (Fig. \ref{fig:MDS_KeggNo}).

In the subsequent differential gene abundance analysis, abundance of 6,104 distinct genes were compared between healthy vs adenoma and healthy vs cancer patients. 
In both cases, there were approximately 130 genes that were less abundant in adenoma and cancer patients, whereas around 600 to 1,000 genes were more abundant. 
Differentially abundant genes that were found in both adenoma and cancer patient groups were selected and used to map to metabolic pathways using the KEGG mapper (see Tables \ref{tab:genes_neg} and \ref{tab:genes_pos}). 

The less and more abundant genes found in adenoma and cancer patients mapped to approximately 20 and 550 genes in the KEGG pathways, respectively. 
Less abundant genes in adenoma and cancer patients were slightly enriched in pathways in metabolism, genetic information processing, signal transduction, transport and catabolism, immune system, and bacterial infectious disease (Table \ref{tab:genes_neg}).
Contrastingly, more abundant genes in adenoma and cancer patients were most enriched in pathways in metabolism, environmental information processing, and cellular proccesses amongst other pathways. 
Though human genome was filtered out through pre-processing, pathways pertaining to human diseases and organismal systems were also enriched in the case of genes more abundant in adenoma and cancer patients.


```{r kegg_genes_neg}
DEgenes_neg <-
  read.csv(file = here(
    "data",
    "metagenome",
    "gene_abundance_results",
    "DEgenes_neg.csv"
  ))
DEgenes_neg$Number.of.Genes <-
  DEgenes_neg$Number.of.Genes %>% replace_na("")
kable(
  DEgenes_neg,
  col.names = c("KEGG Pathway", "", "Number of Genes", "KEGG No."),
  caption = "Genes that are less abundant in adenoma or cancer patients as compared to healthy patients. \\label{tab:genes_neg}"
) %>%
  kable_styling(font_size = 8) %>%
  row_spec(c(1, 17, 20, 22, 24, 26), bold = TRUE)
```


```{r kegg_genes_pos}
DEgenes_pos <-
  read.csv(file = here(
    "data",
    "metagenome",
    "gene_abundance_results",
    "DEgenes_pos.csv"
  ))
DEgenes_pos <- DEgenes_pos %>% replace(., is.na(.), '')
DEgenes_pos <-
  DEgenes_pos %>% filter(Total.Number.of.Genes != "" | KEGG.Pathway != "")
DEgenes_pos$etc <- as.numeric(DEgenes_pos$Total.Number.of.Genes) - as.numeric(DEgenes_pos$Number.of.Genes)
DEgenes_pos$X <- paste(DEgenes_pos$X, if_else(DEgenes_pos$etc == 0, "", ", etc.", ""), sep = '')
DEgenes_pos <- DEgenes_pos %>% select(KEGG.Pathway, X, Total.Number.of.Genes)
kable(
  DEgenes_pos,
  col.names = c("KEGG Pathway", "", "Number of Genes"),
  caption = "Genes that are more abundant in adenoma or cancer patients as compared to healthy patients. \\label{tab:genes_pos}"
) %>%
  kable_styling(font_size = 8) %>%
  row_spec(c(1, 14, 18, 21, 25, 28), bold = TRUE)
```


# Discussion

Based upon the alpha and beta diversity of the OTUs, it is possible that the gut microbiome does not significantly shift in response to the presence of colorectal cancer. Standard alpha and beta diversity metrics do not adequately capture community differences of the bacterial community. This illuminates that examining the 16S sequences of the gut microbiome may not be an adequate tool for identifying a healthy versus a cancerous colon.


# Acknowledgements

Author order was determined by alphabetizing by last name.
Christina Kang-Yun constructed the workflow for preprocessing metagenome data, taxonomic profiling, and gene annotation and prepared the methods and results for these analyses.
\todo{insert contribution statement here}

# References
