---
output:
    pdf_document:
        latex_engine: xelatex
        keep_tex: false
        includes:
          in_header: preamble_report.tex
        toc: true
fontsize: 11pt
geometry: margin=1in
bibliography: bioinf545-group3_report.bib
csl: mbio.csl
link-citations: true
title: "Investigating the microbial community of the human gut in colorectal cancer"
author: "Nicole Bowers, Brittany Hicks, Christina Kang-Yun, Katelyn Polemi, Kelly Sovacool"
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, 
                      fig.path = here::here("figures/"))
```
```{r load}
library(broom)
library(here)
library(knitr)
library(tidyverse)
```

\newpage
\todo{Guidelines: https://drive.google.com/file/d/1YNqYiMJYiVWXiSR9yWUbQTPAoUpVa9Id/view}

# Abstract

# Introduction

# Methods

## Study design and data collection

\todo{study design / data collection based on hannigan paper}
[@hannigan_diagnostic_2017]

The data are available in the NCBI Sequence Read Archive as [PRJNA389927](https://www.ncbi.nlm.nih.gov/bioproject/389927). 
All code used in preparation of this report is available in the following GitHub repository: https://github.com/kelly-sovacool/bioinf545-group3-project.
\todo{make the repo public before submitting}

## 16S rRNA gene processing

## Virome assembly and classification

## Metagenome processing

## Classification modeling

## Metagenome annotation and phylogeny

# Results

\begin{wrapfigure}{R}{0.30\textwidth}
    \begin{center}
        \includegraphics[width=0.29\textwidth]{../figures/rulegraph_metaphlan2.png}  
        \caption{Example}
        \label{fig:example}  
    \end{center}
\end{wrapfigure}

Example of referring to a figure in text (Fig. \ref{fig:example}) 
and including it inline with the text (to be removed before submitting).

# Discussion

# Acknowledgements

Author order was determined by alphabetizing by last name.
\todo{insert contribution statement here}

\newpage
# References
