---
output:
    pdf_document:
        latex_engine: xelatex
        keep_tex: false
        includes:
          in_header: preamble_report.tex
        toc: true
fontsize: 11pt
geometry: margin=1in
bibliography: refs_report.bib
csl: mbio.csl
link-citations: true
title: "Investigating the microbial community of the human gut in colorectal cancer"
author: "Nicole Bowers, Brittany Hicks, Christina Kang-Yun, Katelyn Polemi, Kelly Sovacool"
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, 
                      fig.path = here::here("figures/"))
```
```{r load}
library(broom)
library(here)
library(knitr)
library(tidyverse)
```

\newpage
\todo{Guidelines: https://drive.google.com/file/d/1YNqYiMJYiVWXiSR9yWUbQTPAoUpVa9Id/view}
# Abstract

# Introduction

# Methods

## Study design and data collection

\todo{study design / data collection based on hannigan and zackular papers}
[@hannigan_diagnostic_2017; @Zackular_human_2014]

The data are available in the NCBI Sequence Read Archive as [PRJNA389927](https://www.ncbi.nlm.nih.gov/bioproject/389927). 
All code used in preparation of this report is available in the following GitHub repository: https://github.com/kelly-sovacool/bioinf545-group3-project.
\todo{make the repo public before submitting}

## 16S rRNA gene sequence processing

## Metagenome and virome quality control

Trimmomatic (v.0.39) [@bolger_trimmomatic_2014] was used to remove adapter sequences and low-quality reads in metagenome and virome sample sets. 
The read quality was assessed using FastQC (v.0.11.9) [@andrews_fastqc_2010] before and after adapter trimming to confirm removal of adapters and low quality reads. 
Unpaired reads were dropped using the repair function in BBTools (v.37.62) [@bushnell_bbtools_2018]. 
Then, reads mapping to the human GRCh38 reference genome were removed using the BWA-MEM algorithm (BWA v.0.7.12) [@li_aligning_2013].
The paired unmapped reads were used for taxonomic profiling and gene annotation.

## Virome assembly

## Classification modeling

## Metagenome taxonomic profiling and gene annotation

The human genome-free reads were used to profile the metagenomic taxonomy at the species, genus, family, and phylum levels using MetaPhlAn2 [@truong_metaphlan2_2015]. 
The results were visualized using the R package phyloseq.

The paired reads that did not map to the human genome were aligned to the integrated gene catalog (IGC) [@li_integrated_2014] of 1267 gut microbiome samples consisting of approximately 10 million genes with the BWA MEM algorithm for metagenome annotation [@li_aligning_2013]. 
Annotated genes were extracted from the alignment results using functions geneList and countKegg modified from the MGS-Fast pipeline [@brown_mgs-fast_2019]. 
The difference in gene abundance between healthy, adenoma, and cancer groups were assessed using edgeR. 
The top 500 KEGG numbers of genes that were significantly different between healthy and other groups were selected and the relevant KEGG pathway was determined using the KEGG mapper tool [@kanehisa_kegg_2020].

# Results

\begin{wrapfigure}{R}{0.4\textwidth}
    \begin{center}
        \includegraphics[width=0.35\textwidth]{../figures/rulegraph.png}  
        \caption{Example}
        \label{fig:example}  
    \end{center}
\end{wrapfigure}

Example of referring to a figure in text (Fig. \ref{fig:example}) 
and including it inline with the text.  
\todo{remove example figure before submitting}

# Discussion

# Acknowledgements

Author order was determined by alphabetizing by last name. 
Christina Kang-Yun constructed the workflow for preprocessing metagenome data, taxonomic profiling, and gene annotation and prepared the methods and results for these analyses.
\todo{insert contribution statement here}

# References
